Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-28 review/answer_to_review.md

I was not aware of this work and added a paragraph in the introduction to reference it. It is indeed relevant.

> In the paper, it’s very high the amount of data, for which the archaeological information regarding the type of burial and funerary rite is lacking. At table 1, page 4, we can clearly detect that for 60% of the dates associated to cremation burials we have no information regarding the type of funerary structure. For 37% of the inhumations, no information about the funerary structure is available. Moreover, for 32% of the dataset (559 radiocarbon dates out of 1701) we do not have any kind of information regarding neither the funerary rite nor the burial structure. This lack of data obliges the author to consider carefully the results reached in the paper. It’s relevant to highlight all the biases of the prosed research and how the conclusion can be misleading due to the quality of available data. In particular, when the statistical analyses carried out on data aim to explore in the details the cultural distance among the different regions.

This is absolutely true and I tried to emphasise the limitations of the data in the discussion. The challenges are not confined on the amount of data, but also concern its meaning and origin. The supplementary remarks are extensively long to cover these problems.

I now modified the Figures 3&4 as well as the Table 1 to provide a better overview over the actual amount of data available. I have also added several short comments/sentences in various sections of the paper to better address the small sample size issue.

@@

> Page 3, Column 2, Lines 40-44. The definition of these regions seems rather arbitrary and necessary. Why create such an artificial unit when the spatial coordinates are available?

Please see the reply above.

> Page 3, Column 2, Line 60. “For the remaining 154 dates...”. I assume these dates were excluded as we don’t know whether they are terrestrial or marine?  Also, the paragraph mentions “bones and teeth”, are there any risks of aquatic diet determining a reservoir effect?

These dates were not excluded to avoid further reduction of the amount of data. It is not possible to consider these dating effects on this scale, as it would require to study all radiocarbon dates and their archaeological context individually. I agree that it might be necessary to do so, but this is not an endeavour that can be undertaken without long-term project funding. Marine reservoir effect is only one of the possible problems relevant here in addition to the Old wood effect or the human 14C-residence ("adult") effect. I added a paragraph about this in the *Data quality* section.

@@

> Page 5, Column 1, Line 24. Perhaps worth making this a subsection?

I removed the simulation. See my reply above.

> Page 5, Column 1, Lines 24-60. I have several issues with this model. The idea here is to emulate the observed data so we can build expectations that can be directly compared against empirical patterns. But I do find several issues: 1) How is Ng=100 justified? This seems a somewhat arbitrary number for a parameter that is likely going to play a significant role in the amount of drift in the system? 2) I understand each time-step is 20 years. Why? I assume this is related to generation length but this ought to be discussed; 3) Is the model supposed to reach an equilibrium state? If this is the case I don’t think 70 time-steps and Ng=100 is sufficient for this. If the answer is no, the initial condition of the system shouldn’t be random and perhaps be informed from the empirical data.

I will consider this very important questions in a future publication.

### 2019-03-28 article/02_Materials_and_Methods.Rmd
further work on the text in reaction to the review: small sample size issue

The code for this paper is written in the statistical programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible [@marwickPackagingData2018]. A more in-depth description and discussion of the data preparation steps can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

@@

The dataset used for the analysis in this paper is a subset of the openly available database Radon-B [@kneiselRadonB2013], which contains more than 11000 ^14^C dates from the European Bronze and Early Iron Age (approximately 2300 BC to 500 BC). Radiocarbon dating is a crucial absolute dating technique for this period [@ramseyRadiocarbonDatingRevolutions2008], and Radon-B is a collection of published dates from around 2800 sites located all over Europe (as of February 2019). Beyond the identifying laboratory number, the actual, uncalibrated date and a literature reference, the ^14^C dates in Radon-B are also documented with context information about the name and location of the dated site (coordinates), the sample material and the type and cultural attribution of the sampled context. The variable *feature type* systematically distinguishes between settlements, rock shelters and also different kinds of graves: *Grave (cremation), Grave (flat) inhumation*, i.a.. This information allows to accurately and precisely determine a moment in space and time when a grave of a particular type was created. In theory, the more radiocarbon dates with this information are available, the better becomes our understanding and reconstruction of burial traditions in prehistoric Europe. Unfortunately, the Radon-B dataset has some quality issues mostly due to inconsistent manual data input: Incompleteness and faultiness of meta information, no well-thought-out variable classification for *sample material* and *cultural attribution* and inconsistent multilingualism even within individual entries. These issues though do not inhibit the general data analysis strategy of this paper, as the information about temporal and spatial position as well as *feature type* attribution is mostly complete.

@@

From these time series another set of time series with a measure of cultural distance for each region-region relationship can be derived (Figure \ref{sed_region_matrix}). Cultural distance is defined here as the squared Euclidean distance (SED) between the trait proportions of two different regions at a certain point in time [@neimanStylisticVariationEvolutionary1995]. SED is a sensible metric here, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring [@nakoinzFingerprintingIronAge2014].

>>>>

The code for this paper is written in the statistical programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible [@marwickPackagingData2018]. It is available in a public Repository on GitHub (\url{https://github.com/nevrome/cultrans.bronzeageburials.article2019}) and with a permanent digital object identifier (DOI) on the Open Science Framework Platform (\url{doi}) A more in-depth description and discussion of the data preparation steps can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

@@

The dataset used for the analysis in this paper is a subset of the openly available database Radon-B [@kneiselRadonB2013], which contains more than 11000 ^14^C dates from the European Bronze and Early Iron Age (approximately 2300 BC to 500 BC). Radiocarbon dating is a crucial absolute dating technique for this period [@ramseyRadiocarbonDatingRevolutions2008], and Radon-B is a collection of published dates from around 2800 sites located all over Europe (as of February 2019). Beyond the identifying laboratory number, the actual, uncalibrated date and a literature reference, the ^14^C dates in Radon-B are also documented with context information about the name and location of the dated site (coordinates), the sample material and the type and cultural attribution of the sampled context. The variable *feature type* systematically distinguishes between settlements, rock shelters and also different kinds of graves: *Grave (cremation), Grave (flat) inhumation*, i.a..

This information allows to accurately and precisely determine a moment in space and time when a grave of a particular type was created. In theory, the more radiocarbon dates with this information are available, the better becomes our understanding and reconstruction of burial traditions in prehistoric Europe. The amount of data currently available is relatively small though. The results presented in this paper should therefore be treated with appropriate caution. Several problems of representativity and data meaning are discussed in the sections below.

Furthermore the Radon-B dataset has some quality issues mostly due to inconsistent manual data input: Incompleteness and faultiness of meta information, no well-thought-out variable classification for *sample material* and *cultural attribution* and inconsistent multilingualism even within individual entries. These issues though do not inhibit the general data analysis strategy of this paper, as the information about temporal and spatial position as well as *feature type* attribution is mostly complete.

@@

From these time series another set of time series with a measure of cultural distance for each region-region relationship can be derived (Figure \ref{sed_region_matrix}) -- though with a reasonable amount of caution due to the partially low number of data available to reconstruct the developments in several regions and time slots. Cultural distance is defined here as the squared Euclidean distance (SED) between the trait proportions of two different regions at a certain point in time [@neimanStylisticVariationEvolutionary1995]. SED is a sensible metric here, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring [@nakoinzFingerprintingIronAge2014].
